---
title: "Quality control and differential analysis"
author: "Kimberly Olney, PhD"
date: "2024-12-04"
output: html_document
---

# Set up
```{r setup}
knitr::opts_knit$set(root.dir = ".")
```

```{r libraries}
# Load required libraries
library(tidyverse)
library(ggplot2)
library(pheatmap)
library(limma)
library(edgeR)
library(vsn)
library(ggrepel)
library(xlsx)
library(openxlsx)
library(impute)
library(dplyr)
```

# Input
26ip: contains IP pulldown/eluted sample data
12input: contains input sample data (samples that did not undergo IP pulldown, i.e., the material we started with)
```{r input}
# Load data
raw_ip_data <- read.delim("../raw/MS2_raw_26ip.txt") # MS2_raw_26ip.txt SB28_log2_proteomics.tsv
nsamples <- ncol(raw_ip_data[6:31]) # Number of samples
raw_ip_counts <- raw_ip_data[6:31] # Just the counts 
# Three of the tumor samples didn't graph well. We will consider removing:
# TR_SB28_M_1_elut, TR_SB28_M_2_elut, TR_SB28_M_3_elut
# NOTE if removing be sure to update the columns selected accordingling!! 
raw_ip_counts <- raw_ip_counts[, !colnames(raw_ip_counts) %in% c("TR_SB28_M_1_elut", "TR_SB28_M_2_elut", "TR_SB28_M_3_elut")]

# Remove columns where the name contains "WT_SB28_" or "WT_sham_"
# I.e remove all none TR samples 
raw_ip_counts <- raw_ip_counts[, !grepl("WT_SB28_|WT_sham_", colnames(raw_ip_counts))]

# Select only the first five columns
protein_accession <- raw_ip_data[, c(1:5)]
rownames(raw_ip_counts) <- protein_accession$PG.Genes

# Rename the columns to remove everything after "elut"
colnames(raw_ip_counts) <- gsub("elut.*", "elut", colnames(raw_ip_counts))
metadata <- read.delim("../metadata/SB28_26ip_metadata.tsv")
# Keep only metadata information for samples in the counts matrix
metadata <- subset(metadata, simplified_name %in% colnames(raw_ip_counts))

# Order metadata by matching simplified_name to the column names of raw_ip_counts
#metadata <- metadata[match(colnames(raw_ip_counts), metadata$simplified_name), ]

metadata$treatment <- factor(metadata$treatment, levels = c("sham", "SB28/tumor"))
metadata$sex <- factor(metadata$sex, levels = c("F", "M"))

# Treatment and genotype
metadata$treatment_genotype <- paste(metadata$treatment, metadata$genotype, sep = "_")
#metadata$treatment_genotype <- gsub("SB28/tumor_WT", "SB28_WT", metadata$treatment_genotype)
metadata$treatment_genotype <- gsub("SB28/tumor_TR", "SB28_TR", metadata$treatment_genotype)
metadata$treatment_genotype <- factor(metadata$treatment_genotype, levels = c("sham_TR", "SB28_TR"))
```

# Variables
```{r variables}
SexColors <- c("purple", "orange")
TreatmentColors <- c("lightblue", "black")
TreatmentGenotypeColors <- c("blue", "black") # "lightblue", "blue", "gray", "black"
```

# Counts checks
```{r missing_data}
# Check missing values as a percentage for each protein
missing_vals <- colSums(is.na(raw_ip_data))
missing_vals_percent <- (missing_vals / nrow(raw_ip_data)) * 100
print(missing_vals_percent)

# Visualize missing values distribution
pdf("../TR_only/library/missing_counts_per_sample.pdf", height = 5, width = 7)
ggplot(data.frame(Proteins = rownames(raw_ip_data), MissingPercent = rowMeans(is.na(raw_ip_data))), 
       aes(x = MissingPercent)) +
  geom_histogram(bins = 30, fill = "blue") +
  theme_classic()+  
  labs(title = "Distribution of Missing Values per Protein", x = "Missing Percentage", y = "Frequency")
dev.off()

# Summarize missing data on a per sample 
missing_summary <- colSums(is.na(raw_ip_counts))
print(missing_summary)

# Visualize missing data pattern
missing_df <- data.frame(Sample = colnames(raw_ip_counts), Missing = missing_summary)

# Ensure Sample is a factor with levels in the correct order
missing_df$Sample <- factor(missing_df$Sample, levels = colnames(raw_ip_counts))

pdf("../TR_only/library/missing_counts_per_sample.pdf", height = 5, width = 7)
ggplot(missing_df, aes(x = Sample, y = Missing)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Missing Counts Per Sample", x = "Sample", y = "Count of Missing Proteins") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)  # Rotate X-axis labels to 90 degrees
  )
dev.off()
# min value?
min(raw_ip_counts, na.rm = TRUE) 
```

# Impute missing values
Impute missing values using the k-nearest neighbors (KNN) method.
```{r impute}
data_imputed <- impute.knn(as.matrix(raw_ip_counts))
```

# Correlation between imputed and raw?
```{r}
min_value <- min(raw_ip_counts, na.rm = TRUE) 
raw_ip_counts[is.na(raw_ip_counts)] <- (min_value/2)
imputed_data_check <- as.data.frame(data_imputed$data)

plot(log2(raw_ip_counts$TR_sham_M_1_elut), log2(imputed_data_check$TR_sham_M_1_elut))
plot(log2(raw_ip_counts$TR_SB28_M_4_elut), log2(imputed_data_check$TR_SB28_M_4_elut))
```
```{r}
library(ggplot2)
library(dplyr)

# List the common sample names across both dataframes
common_samples <- intersect(colnames(raw_ip_counts), colnames(imputed_data_check))

# Create a function to generate correlation plots for each sample
plot_correlation <- function(sample_name) {
  # Extract the sample data from both raw_ip_counts and imputed_data_check
  raw_data <- log2(raw_ip_counts[[sample_name]])
  imputed_data <- log2(imputed_data_check[[sample_name]])

  # Create the correlation plot using ggplot
  plot <- ggplot(data = NULL, aes(x = raw_data, y = imputed_data)) +
    geom_point() +
    geom_smooth(method = "lm", col = "blue", linetype = "dashed") +  # Adds a linear regression line
    labs(title = paste("Correlation between", sample_name),
         x = paste("Log2 (", sample_name, " - raw)", sep = ""),
         y = paste("Log2 (", sample_name, " - imputed)", sep = "")) +
    theme_minimal()

  # Save or display the plot
  print(plot)
}

# Loop through each sample and generate a plot
for (sample in common_samples) {
  plot_correlation(sample)
}

```

# vsn mormalization 
```{r vsn}
# Apply VSN normalization
vsn_fit <- vsn2(data_imputed$data)
meanSdPlot(vsn_fit, ranks = TRUE)
counts_matrix <- as.matrix(raw_ip_counts)
data_vsn <- predict(vsn_fit, newdata = raw_ip_counts)
meanSdPlot(data_vsn, ranks= TRUE)
```
The red dots, connected by lines, show the running median of the standard deviation. The aim of these plots is to see whether there is a systematic trend in the standard deviation of the data as a function of overall expression. The assumption that underlies the usefulness of these plots is that most genes are not differentially expressed, so that the running median is a reasonable estimator of the standard deviation of feature level data conditional on the mean. After variance stabilization, this should be approximately a horizontal line. It may have some random fluctuations, but should not show an overall trend. If this is not the case, that usually indicates a data quality problem, or is a consequence of inadequate prior data preprocessing. The rank ordering distributes the data evenly along the x.


# PCA
```{r PCA_post_vsn}
# PCA plot
pca <- prcomp(t(data_vsn), scale. = TRUE)
explained_variance <- pca$sdev^2 / sum(pca$sdev^2) * 100

# PCA plot
pca_data <- as.data.frame(pca$x)
pca_data$treatment <- metadata$treatment_genotype
pca_data$sample <- metadata$simplified_name

# PCA plot visualization
pdf("../TR_only/variance/PCA.pdf", height = 5, width = 7)
ggplot(pca_data, aes(x = PC1, y = PC2, color = treatment)) +
  geom_text(aes(label = sample), vjust = -0.5, hjust = 0.5, size = 3) + # adds sample name
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "PCA of Proteomics Data", 
       x = paste0("PC1 (", round(explained_variance[1], 1), "%)"), 
       y = paste0("PC2 (", round(explained_variance[2], 1), "%)")) +
  scale_color_manual(values = TreatmentGenotypeColors)
dev.off()
```


# Differential expression analysis
```{r DE}
# Check the design matrix
design <- model.matrix(~ 0 + metadata$treatment_genotype)
colnames(design) <- gsub("metadata\\$treatment_genotype", "", colnames(design))

# Fit linear model
fit <- lmFit(data_vsn, design)

# Define contrasts (e.g., Group1 vs Group2)
contrast_matrix <- makeContrasts(
  SB28_TR_vs_sham_TR = SB28_TR- sham_TR,
  levels = design)

# save contrast names
allComparisons <- colnames(contrast_matrix)
allComparisons # check

# Fit contrasts
fit2 <- contrasts.fit(fit, contrast_matrix)
veBayesFit <- eBayes(fit2)

# Extract results
pval <- 0.05
lfc.cutoff <- 0.25

sumTable <- 
  summary(decideTests(
    veBayesFit,  # object
    adjust.method = "BH", # by default the method = "separate"
    p.value = pval,
    lfc = lfc.cutoff  # numeric, minimum absolute log2-fold change required
  ))

print(paste0(" q-value < ", pval,
             " & absolute log2-fold change > ", lfc.cutoff))
sumTable
```


```{r inspect}
test2 <- topTable(
  veBayesFit, 
  coef = "SB28_TR_vs_sham_TR",  
  n = Inf, 
  p.value = 1,
  lfc = 0, 
  sort.by = "P", 
 # genelist = protein_accession, 
  confint = TRUE # column of confidence interval 
    )

subset(test2, rownames(test2) == "Mapt")
```


### Output DEG tables
```{r DGE_output}
coef <- 1

for (i in allComparisons) {
  vTopTableAll <- topTable(
    veBayesFit, 
    coef = coef,  
    n = Inf, 
    p.value = 1,
    lfc = 0, 
    sort.by = "P", 
    confint = TRUE 
    )
  vTopTableAll$PG.Genes <- rownames(vTopTableAll)
  df <- merge(vTopTableAll, protein_accession, by = "PG.Genes")
  saveRDS(df, file = paste0("../rObjects/", 
                   i,"_gene_table.rds"))
  
  path <- paste0("../TR_only/differential/", 
  i, "_gene_DEGs_q1.00.txt", sep = "") 
  write.table(
    df,
    path,
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )
  
  vTopTable1 <- subset(df, (adj.P.Val < 0.05 & logFC > 0.25) | (adj.P.Val < 0.05 & logFC < -0.25))
  path <- paste0("../TR_only/differential/", 
  i, "_gene_DEGs_q0.05_logFC0.25.txt", sep = "") 
  write.table(
    vTopTable1,
    path,
    sep = "\t",
    row.names = FALSE,
    quote = FALSE
  )
  coef <- coef + 1  # increment 
}
remove(coef, df, vTopTable1, vTopTableAll)
```

# Volcano plots
Add label for turbo ID 
```{r volcano, warning=FALSE}
q_value_cutoff = 0.05
logFC_cutoff = 0.25
for (i in allComparisons) {
  group1_vs_group2 <-
    read.delim(
      paste0("../TR_only/differential/",
        i,
        "_gene_DEGs_q1.00.txt"
      )
    )
  color_values <- vector()
  max <- nrow(group1_vs_group2)
  for (row in 1:max) {
    if (group1_vs_group2$adj.P.Val[row] < 0.05) {
      if (group1_vs_group2$logFC [row] > 0.25) {
        color_values <- c(color_values, 1)
      }
      else if (group1_vs_group2$logFC[row] < -0.25) {
        color_values <- c(color_values, 2)
      }
      else {
        color_values <- c(color_values, 3)
      }
    }
    else{
      color_values <- c(color_values, 3)
    }
  }
  group1_vs_group2$color_adjpval_0.05 <- factor(color_values)
  data <- group1_vs_group2
  # plot only if there are DEGs with p_val_adj < 0.05
  num <- subset(data, (adj.P.Val < 0.05 & logFC < -0.25)  | (adj.P.Val < 0.05 & logFC > 0.25 ))
  num <- nrow(num)
  if (num != 0) {
    up <- data[data$color_adjpval_0.05 == 1,]
    up10 <- up[1:10,]
    upFold <- subset(up, logFC > 0)
    upFold <- upFold[!(upFold$PG.Genes %in% up10$PG.Genes),]
    down <- data[data$color_adjpval_0.05 == 2,]
    down10 <- down[1:10,]
    downFold <- subset(down, logFC < -0)
    downFold <- downFold[!(downFold$PG.Genes %in% down10$PG.Genes),]
    if (!1 %in% unique(data$color_adjpval_0.05)) {
      my_colors <- c("blue", "gray")
    } else if (!2 %in% unique(data$color_adjpval_0.05)) {
      my_colors <- c("red", "gray")
    } else if (!1 %in% unique(data$color_adjpval_0.05) &&
               !2 %in% unique(data$color_adjpval_0.05)) {
      my_colors <- c("gray")
    } else {
      my_colors <- c("red", "blue", "gray")
    }
    hadjpval <- (-log10(max(data$P.Value[data$adj.P.Val < 0.05],
                            na.rm = TRUE)))
    negFC <- c(-0.25)
    posFC <- c(0.25) 
    
    title_main <- gsub("_vs_", " versus ", gsub("_", " ", i)) # format group name with parentheses
    title_sub <- paste("q-value <", q_value_cutoff, "& |log2FC| >", logFC_cutoff)
    
    i <- gsub("vs", " versus ", i)
    p <-
      ggplot(data = data,
             aes(
               x = logFC,
               y = -log10(P.Value),
               color = color_adjpval_0.05
             )) +
      geom_point(alpha = 0.8, size = 1) +
      theme_bw() +
      theme(legend.position = "none") +
      scale_color_manual(values = my_colors) +
      labs(
        title = title_main,
        subtitle = title_sub,
        x = expression(log[2](FC)),
        y = expression(-log[10] ~ "(" ~ italic("p") ~ "-value)")
      )  +
      geom_hline(yintercept = hadjpval,
                 #  horizontal line
                 colour = "#000000",
                 linetype = "dashed") +
      geom_vline(xintercept = negFC,
                 #  horizontal line
                 colour = "#000000",
                 linetype = "dashed") +
      geom_vline(xintercept = posFC,
                 #  horizontal line
                 colour = "#000000",
                 linetype = "dashed") +
      geom_text_repel(
        data = up10,
        aes(
          x = logFC,
          y = -log10(P.Value),
          label = PG.Genes
        ),
        color = "maroon",
        fontface = "italic",
        size = 3,
        max.overlaps = getOption("ggrepel.max.overlaps", default = 5)
      ) +
      geom_text_repel(
        data = upFold,
        aes(
          x = logFC,
          y = -log10(P.Value),
          label = PG.Genes
        ),
        color = "maroon",
        fontface = "italic",
        size = 3,
        max.overlaps = getOption("ggrepel.max.overlaps", default = 5)
      ) +
      geom_text_repel(
        data = down10,
        aes(
          x = logFC,
          y = -log10(P.Value),
          label = PG.Genes
        ),
        color = "navyblue",
        fontface = "italic",
        size = 3,
        max.overlaps = getOption("ggrepel.max.overlaps", default = 5)
      ) +
      geom_text_repel(
        data = downFold,
        aes(
          x = logFC,
          y = -log10(P.Value),
          label = PG.Genes
        ),
        color = "navyblue",
        fontface = "italic",
        size = 3,
        max.overlaps = getOption("ggrepel.max.overlaps", default = 5)
      )
    p
    i <- gsub(" versus ", "vs", i)
    # save
    path <-
      paste0(
        "../TR_only/volcano/",
        i,
        "_volcano_q0.05_logFC0.25"
      )
    pdf(paste0(path, ".pdf"),
        height = 5,
        width = 5)
    print(p)
    dev.off()
    
    print(paste("i =", i))
  }
} 
```

```{r cleanup}
# clean up
remove(up, up10, upFold, group1_vs_group2, downFold, down10, data, p)
```

# Make excel table
```{r excel}
# first read in the files
for (i in allComparisons) {
  filepath <- paste0("../TR_only/differential/", 
  i, "_gene_DEGs_q1.00.txt", sep = "") 
  assign(paste0(i),
         read.delim(
           filepath,
           header = TRUE,
           sep = "\t",
           stringsAsFactors = FALSE
         ))
}
allComparisons

list_of_datasets <- list(
"SB28_TR_vs_sham_TR" = SB28_TR_vs_sham_TR        
)

write.xlsx(list_of_datasets, file = paste0(
    "../TR_only/differential/DEGs_q1.00.xlsx"))
```

clean up
```{r cleanup_excel}
remove(
SB28_TR_vs_sham_TR
)
```

# Excel table of significant DEGs
```{r}
# first read in the files
for (i in allComparisons) {
  filepath <- paste0("../TR_only/differential/", 
  i, "_gene_DEGs_q0.05_logFC0.25.txt", sep = ""
  )
  assign(paste0(i),
         tryCatch(
         read.delim(
           filepath,
           header = TRUE,
           sep = "\t",
           stringsAsFactors = FALSE
         ), error=function(e) NULL))
}

list_of_datasets <- list(
"SB28_TR_vs_sham_TR" = SB28_TR_vs_sham_TR)
write.xlsx(list_of_datasets, file = paste0("../TR_only/differential/DEGs_q0.05_logFC0.25.xlsx"))
```

```{r}
# Create empty lists to store up-regulated and down-regulated genes dataframes
up_regulated_genes <- list()
down_regulated_genes <- list()

# Iterate over each dataframe in list_of_datasets
for (comparison_name in names(list_of_datasets)) {
    # Filter for up-regulated genes (logFC > 0)
    up_regulated_genes[[comparison_name]] <- list_of_datasets[[comparison_name]][list_of_datasets[[comparison_name]]$logFC > 0, ]
    
    # Filter for down-regulated genes (logFC < 0)
    down_regulated_genes[[comparison_name]] <- list_of_datasets[[comparison_name]][list_of_datasets[[comparison_name]]$logFC < 0, ]
}

# Write up-regulated genes to an Excel file
up_regulated_file <- paste0("../TR_only/differential/upregulated_proteins.xlsx")
write.xlsx(up_regulated_genes, file = up_regulated_file)

# Write down-regulated genes to an Excel file
down_regulated_file <- paste0("../TR_only/differential/downregulated_proteins.xlsx")
write.xlsx(down_regulated_genes, file = down_regulated_file)
```



